Investigating soundscapes perception through acoustic scenes simulation.

### 2017-09-01 lafaySimulatedScenePerception.tex
improve simulator sec and planning sec

% L'objectif est d'étudier les influences spécifiques des différentes sources sonores qui composent les environnements urbains sur la perception de l'agrément, en utilisant la simulation. Pour ce faire, nous planifions nos deux premières expériences (\cf~Figure~\ref{fig:xp1_2}):
%
% \begin{itemize}
% \item \emph{expérience 1.a de simulation}: au cours de cette expérience, les sujets doivent simuler des environnements sonores urbains, en utilisant l'outil de simulation \emph{Simscene} (\cf~Section~\ref{sec:simscene}). Chacun compose deux environnements sonores, le premier idéal/agréable et le deuxième non-idéal/désagréable.
%
% \item \emph{expérience 1.b d'évaluation}: à l'issue de la simulation, nous n'avons, de fait, qu'une connaissance binaire des propriétés affectives des scènes simulées: idéale (i) et non-idéale (ni). Cette seconde étape a pour but d'affiner notre connaissance sur l'agrément. Pour ce faire, nous demandons à un deuxième groupe de sujets d'évaluer, à partir d'une échelle sémantique, l'agrément de chacune des scènes simulées. L'expérience d'évaluation sert deux buts:
%
% \begin{enumerate}
% \item évaluer l'influence respective des différentes sources sur l'agrément, pour chaque type d'environnement (i ou ni);
% \item détecter la présence de cas extrêmes ou ambigus (\emph{outlier}) dans les scènes simulées. Pour le reste de notre étude, les qualités hédoniques imposées (i et ni) servent de référence, de vérité terrain. Il nous faut donc garantir qu'il n'y ait pas d’ambiguïté entre les cas extrêmes des i- et ni-scènes, \ie~que la note d'agrément la plus basse des i-scènes reste supérieure à la note la plus haute des ni-scènes.
% \end{enumerate}
%
% \end{itemize}
%
% Notre analyse s'appuie sur les données produites par les deux expériences.

Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness. For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}. Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on pleasantness. For that, a second group of participants is asked to evaluate the pleasantness of each simulated scene, on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
\item to detect the presence of \emph{outliers} or ambiguous samples. For the remainder of the study, the defined hedonic properties (i and ni) are used as references. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of perceived pleasantness.
\end{enumerate}
\end{itemize}

Our analysis is based on the data collected by these two experiments: 1.a and 1.b.

@@

@@

% Les sujets doivent simuler deux environnements sonores urbains, chacune des scènes devant durer 1 minute. Ils doivent se conformer aux consignes suivantes:
%
% \begin{itemize}
% \item première simulation: simuler un paysage sonore \textbf{urbain plausible} qui, selon vous, est idéal (où vous aimeriez vivre);
% \item deuxième simulation: simuler un paysage sonore \textbf{urbain plausible} qui, selon vous, est non-idéal (où vous n'aimeriez pas vivre).
% \end{itemize}
%
% Tous commencent par simuler l'environnement idéal. Ils ne prennent connaissance de la deuxième consigne qu'à la fin de la première simulation.
%
% Les sujets sont totalement libres dans le choix des sons, et des paramètres. Ils doivent cependant se soumettre à deux contraintes:
%
% \begin{itemize}
% \item le sujet doit prendre le point de vue d’un auditeur fixe;
% \item le paysage sonore doit être réaliste, au sens de physiquement plausible. Le sujet peut très bien réunir 10 chiens dans un paysage sonore. Il ne peut en introduire 1 qui aboierait toutes les 10 millisecondes.
% \end{itemize}
%
% Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.

The participants have to simulate two urban sound environments of one minute each. They follow the instructions below:
\begin{itemize}
\item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
\item second simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
\end{itemize}

All the participants start by designing the ideal environment; they read the second set of instruction at the end of the first experiment. Participants are completely free of their choices concerning sounds and synthesis parameters. They must nevertheless respect the two following constraints :
\begin{itemize}
\item the listening point of view is the one of a fixed listener;
\item the soundscape must be realistic, i.e. physically plausible. For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
\end{itemize}

@@

% Chaque processus de simulation comprend deux parties :
%
% \begin{enumerate}
% \item la réalisation de la simulation: cette étape se décompose en trois temps:
% \begin{itemize}
% \item sélectionner les classes de sons
% \item nommer les classes de sons sélectionnées
% \item paramétrer les pistes (\cf~Section~\ref{sec:simscene_parametre}) relatives aux classes de sons sélectionnées
% \end{itemize}
% \item la production d'un commentaire libre du paysage sonore simulé
% \end{enumerate}
%
% En complément, et une fois les deux scènes sonores réalisées, les sujets sont invités à:
%
% \begin{itemize}
% \item indiquer les sources sonores qu'ils voulaient mettre, mais qu'ils n'ont pas trouvées;
% \item commenter l’ergonomie du logiciel de simulation;
% \item commenter l’ergonomie de l'interface de sélection.
% \end{itemize}
%
% Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données. L'expérience est prévue pour durer 2h30.

Each simulation process contains two parts:
\begin{enumerate}
\item composition of the simulation: this step is split in three steps:
\begin{itemize}
  \item  select the class of sounds
	\item  give a name to the selected class of sounds
	\item  set the parameters of the tracks related to the selected class of sounds, see Section~\ref{sec:simscene_parametre}.
\end{itemize}
\item production of a free verbalization of the composed soundscape
\end{enumerate}

In addition, once the two sound scenes are done, participants are invited to
\begin{itemize}
\item  point out the sound sources that were missing for the composition
\item  give a comment about the ergonomics of the simulation environment
\item   give a comment about the ergonomics of the selection tool
\end{itemize}

Before starting the first simulation, a 20-minute tutorial is given to the participants in order to familiarize them with the simulation environment and the sound database. The experiment is planned to last about two hours and a half, including breaks that could be taken by the participants.

@@

% Tous les sujets passent l'expérience sur des machines identiques. L'audio est diffusé en diotique, par le biais de casques audio. Pendant le tutoriel, les sujets doivent ajuster le niveau sonore à un volume confortable. Ils ne peuvent le modifier par la suite.
%
% Tous les sujets réalisent l'expérience simultanément. Ils sont répartis de manière égale dans trois pièces identiques, toutes possédant un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
%
% Trois expérimentateurs, un dans chaque pièce, sont présents durant la totalité de l'expérience, afin de contrôler le bon déroulement de cette dernière, et de répondre aux éventuelles questions des sujets.

All the participants do the experiment on desktop computers with the same hardware and software configurations. The audio files are played in diotic conditions by headphones. During the tutorial, participants are asked to adjust the sound level in order to be comfortable. Once set, they are not allowed to modify it in the remaining of the experiment.
All the participants do the experiment at the same time. They are equally split in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
Three experimenters (one in each room) are present during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.

@@

% 44 étudiants (14 femmes) de L’École Centrale de Nantes ont participé à l'expérience. Ils ont tous sensiblement le même âge (moyenne: 21.6, écart-type: 2). Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
%
% Sur les 44 sujets, 40 réalisent l'expérience avec succès, produisant au final 80 scènes sonores simulées, dont 40 scènes idéales, et 40 scènes non idéales. 4 sont éliminés pour non respect ou incompréhension des consignes, d'une part, et dépassement du temps, d'autre part.

44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in downtown Nantes since 2 years or more.
Among the 44 participants, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). the 4 other participants were excluded from the process due to a lack of respect or understanding of the instructions, or for exceeding the maximum duration of the experiment.

@@

% En raison de contraintes temporelles, les sujets n'évaluent que des séquences de 30 secondes des scènes simulées, chacune de ces séquences commençant à la seconde 15, et finissant à le seconde 45 de la scène d'une durée initiale de 1 minute.
%
% L'évaluation s'effectue sur une échelle sémantique bipolaire de 7 points, allant de -3 (non-idéale/très désagréable) à +3 (idéale/très agréable). Avant de noter une scène, les sujets doivent obligatoirement écouter les 20 premières secondes de cette dernière. Après la notation, ils sont libres de passer à la scène suivante.
%
% Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire. Les 10 premières scènes permettent au sujet de calibrer ses notes. Elles sont obligatoirement composées de 5 scènes idéales, et de 5 scènes non-idéales. Ces 10 premières scènes sont rejouées à la fin de l'expérience, et seules les notes données à la deuxième occurrence sont prises en compte.
%
% L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.

Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes composed in Experiment 1.a (from sec. 15 to sec. 45).
The assessment is done with a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli. After the evaluation, they are free to switch to the next scene.
For each participant, sound scenes are played in a random order. The first 10 scenes allow the participants to calibrate their scores; as they are always composed of 5 ideal scenes and 5 non ideal scenes. These first 10 scenes are played back again at the end of the experiment and the last evaluations are the ones that are only taken into account in the results.
The experiment is planned to last 30 minutes. The participants do not know anything beforehand about the nature of the sound scenes.

@@

% Tous les sujets passent l'expérience sur des machines identiques. L'audio est diffusé en monophonie, par le biais de casques audio semi-ouverts \emph{Beyer-Dynamic DT 990 Pro}. Toutes les scènes sonores ont été re-simulées sur la base des partitions obtenues lors de l'expérience de simulation. Le niveau sonore de sortie est identique pour tous les sujets.
%
% Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
%
% Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de celle-ci, et de répondre aux éventuelles questions des sujets.

All the participants do the experiment on computers with equal hardware and software configurations. The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli are the scenes obtained by the simulation experiment of Exp. 1.a. The output sound level is the same for all the participants.
All the participants do the experiment simultaneously. They are not allowed to talk to each other during the experiment.
An experimenter is present during the whole duration of the experiment in order to assist participants and possibly answer questions.

@@

% 10 étudiants (2 femmes) de L’École Centrale de Nantes ont participé à l'expérience. Aucun d'entre eux n'a réalisé l'expérience de simulation. Tous les sujets ont sensiblement le même âge (moyenne: 23.1, écart-type: 1.8). Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
%
% Tous les sujets ont réalisé l'expérience avec succès.

10 students (2 females - mean age 23.1, SD = 1.8) from Ecole Centrale de Nantes took part to the experiment. All are living in downtown Nantes for 2 years or more. None of them took part to the previous simulation experiment (Exp. 1.a).
All the participants succeeded in doing the experiment.

\subsection{Data and analysis methods}
\label{sec:xp1_dataAna}

% A chaque scène est associé un groupe de descripteurs. C'est sur ces descripteurs que nous pratiquons l'analyse. Un résumé de ces derniers, ainsi que des acronymes les désignant, est présenté dans le Tableau~\ref{tab:acronyme}. Afin de rester cohérent avec l’expérience 1.b d'évaluation, les descripteurs ne sont pas calculés sur la durée totale des scènes, mais sur la version réduite de 30 secondes (\cf~Section~\ref{sec:xp1b_plan}).
%
% Pour chaque scène sonore, trois types de descripteurs sont considérés:
%
% \begin{itemize}
% \item \emph{perceptif}: il s'agit de l'agrément perçu de la scène simulée, évalué sur une échelle sémantique de 7 points. Nous notons $\mathcal{A}_{scene}$ l'agrément moyen d'une scène, obtenu en moyennant les notes de tous les sujets. De même, nous notons $\mathcal{A}_{sujet}$ l'agrément par sujet, en moyennant l'ensemble des notes de chacun. Compte tenu du faible nombre de sujets, nous faisons le choix, dans cette étude, de ne pas normaliser les notes d'agrément;
% \item \emph{sémantique}: il s'agit d'un vecteur booléen noté $S=(x_1,x_2,\ldots,x_n)$, indiquant les classes de sons présentes dans la scène. Chaque point $x$ de ce vecteur correspond à une classe de sons particulière: $x=1$ si la classe est présente dans la scène, et $x=0$ autrement. La dimension $n$ des vecteurs dépend du niveau d'abstraction considéré, \eg~pour le niveau d'abstraction 1, qui comprend $44$ classes de sons, cette dimension sera de $n=44$.
% \item \emph{structurel}: il s'agit du \emph{niveau Sonore}. Pour représenter le niveau sonore, nous nous inspirons de la mesure $L_{Aeq}$. Dans notre cas, il s'agit d'un scalaire calculé sur le signal en volts, et non en pression, et donné en décibels, en prenant un référentiel de 1 Volt. Le niveau est obtenu en calculant, toutes les secondes, la moyenne quadratique du signal, et en moyennant sur la durée de la scène. Un filtrage de type A est opéré avant le calcul des moyennes quadratiques. Nous notons $L$, $L(E)$ et $L(T)$ les niveaux calculés en tenant compte respectivement de l'ensemble des samples, des samples d'événements et de textures.
% \end{itemize}

A set of features is attached to each sound scene. The analysis is done on the basis of these features. A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}. In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.

@@

\begin{itemize}
\item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of participants in Exp. 1.b, we choose not to normalize the pleasantness scores.
\item \emph{semantic}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for analysis. For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
\item \emph{structural}: features related to \emph{sound level}. To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
\end{itemize}

@@

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                         & Acronyms              \\
\hline
Level                        & $L$                    \\
Level (events)            & $L(E)$                 \\
Level (textures)              & $L(T)$                 \\
Average pleasantness (per scene)     & $\mathcal{A}_{scene}$  \\
Average pleasantness (per subject)     & $\mathcal{A}_{sujet}$  \\
                               &                        \\
ideal/pleasant                & i                      \\
non-ideal/unpleasant         & ni                     \\
scene ideal/pleasant          & i-scene                \\
scene non-ideal/unpleasant   & ni-scene               \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Acronyms of features used in the analysis of the sensory experiments.}
\label{tab:acronyme}
\end{table}

>>>>

@@

% L'objectif est d'étudier les influences spécifiques des différentes sources sonores qui composent les environnements urbains sur la perception de l'agrément, en utilisant la simulation. Pour ce faire, nous planifions nos deux premières expériences (\cf~Figure~\ref{fig:xp1_2}):

Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness. For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):

% \begin{itemize}
% \item \emph{expérience 1.a de simulation}: au cours de cette expérience, les sujets doivent simuler des environnements sonores urbains, en utilisant l'outil de simulation \emph{Simscene} (\cf~Section~\ref{sec:simscene}). Chacun compose deux environnements sonores, le premier idéal/agréable et le deuxième non-idéal/désagréable.
% \item \emph{expérience 1.b d'évaluation}: à l'issue de la simulation, nous n'avons, de fait, qu'une connaissance binaire des propriétés affectives des scènes simulées: idéale (i) et non-idéale (ni). Cette seconde étape a pour but d'affiner notre connaissance sur l'agrément. Pour ce faire, nous demandons à un deuxième groupe de sujets d'évaluer, à partir d'une échelle sémantique, l'agrément de chacune des scènes simulées. L'expérience d'évaluation sert deux buts:
% \begin{enumerate}
% \item évaluer l'influence respective des différentes sources sur l'agrément, pour chaque type d'environnement (i ou ni);
% \item détecter la présence de cas extrêmes ou ambigus (\emph{outlier}) dans les scènes simulées. Pour le reste de notre étude, les qualités hédoniques imposées (i et ni) servent de référence, de vérité terrain. Il nous faut donc garantir qu'il n'y ait pas d’ambiguïté entre les cas extrêmes des i- et ni-scènes, \ie~que la note d'agrément la plus basse des i-scènes reste supérieure à la note la plus haute des ni-scènes.
% \end{enumerate}
% \end{itemize}

\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to simulate urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}. Each of them has to create two sound environments: one ideal / pleasant and the other one, non ideal / unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes. For that, a second group of participants is asked to evaluate the pleasantness of each scene, on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
\item to detect the presence of \emph{outliers} or ambiguous scenes. Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as groundtruth. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the less pleasant i-scene remains above the more pleasant ni-scene.
\end{enumerate}
\end{itemize}

% Notre analyse s'appuie sur les données produites par les deux expériences.

The data collected by these two experiments (1.a and 1.b) are analyzed conjointly.

@@

% Les sujets doivent simuler deux environnements sonores urbains, chacune des scènes devant durer 1 minute. Ils doivent se conformer aux consignes suivantes:

The participants have to simulate two urban sound environments of one minute each. They follow the instructions below:

% \begin{itemize}
% \item première simulation: simuler un paysage sonore \textbf{urbain plausible} qui, selon vous, est idéal (où vous aimeriez vivre);
% \item deuxième simulation: simuler un paysage sonore \textbf{urbain plausible} qui, selon vous, est non-idéal (où vous n'aimeriez pas vivre).
% \end{itemize}

\begin{itemize}
\item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
\item second simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
\end{itemize}

% Tous commencent par simuler l'environnement idéal. Ils ne prennent connaissance de la deuxième consigne qu'à la fin de la première simulation.

% Les sujets sont totalement libres dans le choix des sons, et des paramètres. Ils doivent cependant se soumettre à deux contraintes:

All the participants start by designing the ideal environment; they read the second set of instruction at the end of the first experiment. Participants are completely free of their choices concerning sounds and synthesis parameters. They must nevertheless respect the two following constraints :

% \begin{itemize}
% \item le sujet doit prendre le point de vue d’un auditeur fixe;
% \item le paysage sonore doit être réaliste, au sens de physiquement plausible. Le sujet peut très bien réunir 10 chiens dans un paysage sonore. Il ne peut en introduire 1 qui aboierait toutes les 10 millisecondes.
% \end{itemize}

\begin{itemize}
\item the listening point of view is the one of a fixed listener;
\item the soundscape must be realistic, \ie~physically plausible. For instance, participants are free to insert ten dogs in the soundscape but they cannot insert one dog barking every 10 milliseconds.
\end{itemize}

% Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.

@@

% Chaque processus de simulation comprend deux parties :
%
% \begin{enumerate}
% \item la réalisation de la simulation: cette étape se décompose en trois temps:
% \begin{itemize}
% \item sélectionner les classes de sons
% \item nommer les classes de sons sélectionnées
% \item paramétrer les pistes (\cf~Section~\ref{sec:simscene_parametre}) relatives aux classes de sons sélectionnées
% \end{itemize}
% \item la production d'un commentaire libre du paysage sonore simulé
% \end{enumerate}

Each simulation process comprises two parts:

\begin{enumerate}
\item the simulation: this step is split in three steps:
\begin{itemize}
  \item  select the class of sounds
	\item  give a name to the selected class of sounds
	\item  set the parameters of the tracks related to the selected class of sounds, see Section~\ref{sec:simscene_parametre}.
\end{itemize}
\item the production of a free comment of the composed soundscape
\end{enumerate}

% En complément, et une fois les deux scènes sonores réalisées, les sujets sont invités à:
%
% \begin{itemize}
% \item indiquer les sources sonores qu'ils voulaient mettre, mais qu'ils n'ont pas trouvées;
% \item commenter l’ergonomie du logiciel de simulation;
% \item commenter l’ergonomie de l'interface de sélection.
% \end{itemize}

In addition, once the two sound scenes are done, participants are invited to:

\begin{itemize}
\item  point out the sound sources that were missing for the composition
\item  give a comment about the ergonomics of the simulation environment
\item   give a comment about the ergonomics of the selection tool
\end{itemize}

% Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données. L'expérience est prévue pour durer 2h30.

Before starting the first simulation, a 20-minute tutorial is given to the participants in order to familiarize them with the simulation interface and the sound database. The experiment is planned to last about two hours and a half, including breaks that could be taken by the participants.

@@

% Tous les sujets passent l'expérience sur des machines identiques. L'audio est diffusé en diotique, par le biais de casques audio. Pendant le tutoriel, les sujets doivent ajuster le niveau sonore à un volume confortable. Ils ne peuvent le modifier par la suite.

All the participants do the experiment on desktop computers with the same hardware and software configurations. The audio files are played in diotic conditions by headphones. During the tutorial, participants are asked to adjust the global sound volume to a comfortable level. Once set, they are not allowed to modify it in the remaining of the experiment.

% Tous les sujets réalisent l'expérience simultanément. Ils sont répartis de manière égale dans trois pièces identiques, toutes possédant un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.

All the participants do the experiment at the same time. They are equally split in three identical quiet rooms. They are not allowed to talk to each other during the experiment.

% Trois expérimentateurs, un dans chaque pièce, sont présents durant la totalité de l'expérience, afin de contrôler le bon déroulement de cette dernière, et de répondre aux éventuelles questions des sujets.

Three experimenters (one in each room) are present during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.

@@

% 44 étudiants (14 femmes) de L’École Centrale de Nantes ont participé à l'expérience. Ils ont tous sensiblement le même âge (moyenne: 21.6, écart-type: 2). Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.

44 students (14 females; averaging 21.6 years of age, s.d. of 2.0 years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment. They are all living in downtown Nantes (a french city) since 2 years or more.

% Sur les 44 sujets, 40 réalisent l'expérience avec succès, produisant au final 80 scènes sonores simulées, dont 40 scènes idéales, et 40 scènes non idéales. 4 sont éliminés pour non respect ou incompréhension des consignes, d'une part, et dépassement du temps, d'autre part.

Among the 44 participants, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). the 4 other participants were excluded from the process due to a lack of respect or understanding of the instructions, or for exceeding the maximum duration of the experiment.

@@

% En raison de contraintes temporelles, les sujets n'évaluent que des séquences de 30 secondes des scènes simulées, chacune de ces séquences commençant à la seconde 15, et finissant à le seconde 45 de la scène d'une durée initiale de 1 minute.

Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).

% L'évaluation s'effectue sur une échelle sémantique bipolaire de 7 points, allant de -3 (non-idéale/très désagréable) à +3 (idéale/très agréable). Avant de noter une scène, les sujets doivent obligatoirement écouter les 20 premières secondes de cette dernière. Après la notation, ils sont libres de passer à la scène suivante.

The assessment is done with a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli. After the evaluation, they are free to switch to the next scene.

% Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire. Les 10 premières scènes permettent au sujet de calibrer ses notes. Elles sont obligatoirement composées de 5 scènes idéales, et de 5 scènes non-idéales. Ces 10 premières scènes sont rejouées à la fin de l'expérience, et seules les notes données à la deuxième occurrence sont prises en compte.

For each participant, sound scenes are played in a random order. The first 10 scenes allow the participants to calibrate their scores; as they are always composed of 5 ideal scenes and 5 non ideal scenes. These first 10 scenes are played back again at the end of the experiment and the last evaluations are the ones that are only taken into account in the results.

% L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.

The experiment is planned to last 30 minutes. The participants do not know anything beforehand about the nature of the sound scenes.

@@

% Tous les sujets passent l'expérience sur des machines identiques. L'audio est diffusé en monophonie, par le biais de casques audio semi-ouverts \emph{Beyer-Dynamic DT 990 Pro}. Toutes les scènes sonores ont été re-simulées sur la base des partitions obtenues lors de l'expérience de simulation. Le niveau sonore de sortie est identique pour tous les sujets.

All the participants do the experiment on computers with equal hardware and software configurations. The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli are the scenes obtained by the simulation experiment of experiment 1.a. The output sound level is the same for all the participants.

% Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.

All the participants do the experiment simultaneously. They are not allowed to talk to each other during the experiment.

% Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de celle-ci, et de répondre aux éventuelles questions des sujets.

An experimenter is present during the whole duration of the experiment in order to assist participants and possibly answer questions.

@@

% 10 étudiants (2 femmes) de L’École Centrale de Nantes ont participé à l'expérience. Aucun d'entre eux n'a réalisé l'expérience de simulation. Tous les sujets ont sensiblement le même âge (moyenne: 23.1, écart-type: 1.8). Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.

10 students (2 females; averaging 23.1 years of age, s.d. of 1.8 years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment. They are living in downtown Nantes (a french city) since 2 years or more. None of them took part to the previous simulation experiment (Exp. 1.a).

% Tous les sujets ont réalisé l'expérience avec succès.

All the participants succeeded in doing the experiment.

\subsection{Data and statistical analysis}
\label{sec:xp1_dataAna}

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                         & Acronyms              \\
\hline
sound Level                        & $L$                    \\
sound Level (events)            & $L(E)$                 \\
sound Level (textures)              & $L(T)$                 \\
Average pleasantness (per scene)     & $\mathcal{A}_{scene}$  \\
Average pleasantness (per subject)     & $\mathcal{A}_{sujet}$  \\
                               &                        \\
ideal/pleasant                & i                      \\
non-ideal/unpleasant         & ni                     \\
scene ideal/pleasant          & i-scene                \\
scene non-ideal/unpleasant   & ni-scene               \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Acronyms of features used in the analysis of the sensory experiments.}
\label{tab:acronyme}
\end{table}

% A chaque scène est associé un groupe de descripteurs. C'est sur ces descripteurs que nous pratiquons l'analyse. Un résumé de ces derniers, ainsi que des acronymes les désignant, est présenté dans le Tableau~\ref{tab:acronyme}. Afin de rester cohérent avec l’expérience 1.b d'évaluation, les descripteurs ne sont pas calculés sur la durée totale des scènes, mais sur la version réduite de 30 secondes (\cf~Section~\ref{sec:xp1b_plan}).

A set of features is attached to each sound scene. The analysis is done on the basis of these features. A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}. In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section~\ref{sec:xp1b_plan}.

% Pour chaque scène sonore, trois types de descripteurs sont considérés:

@@

% \begin{itemize}
% \item \emph{perceptif}: il s'agit de l'agrément perçu de la scène simulée, évalué sur une échelle sémantique de 7 points. Nous notons $\mathcal{A}_{scene}$ l'agrément moyen d'une scène, obtenu en moyennant les notes de tous les sujets. De même, nous notons $\mathcal{A}_{sujet}$ l'agrément par sujet, en moyennant l'ensemble des notes de chacun. Compte tenu du faible nombre de sujets, nous faisons le choix, dans cette étude, de ne pas normaliser les notes d'agrément;
% \item \emph{sémantique}: il s'agit d'un vecteur booléen noté $S=(x_1,x_2,\ldots,x_n)$, indiquant les classes de sons présentes dans la scène. Chaque point $x$ de ce vecteur correspond à une classe de sons particulière: $x=1$ si la classe est présente dans la scène, et $x=0$ autrement. La dimension $n$ des vecteurs dépend du niveau d'abstraction considéré, \eg~pour le niveau d'abstraction 1, qui comprend $44$ classes de sons, cette dimension sera de $n=44$.
% \item \emph{structurel}: il s'agit du \emph{niveau Sonore}. Pour représenter le niveau sonore, nous nous inspirons de la mesure $L_{Aeq}$. Dans notre cas, il s'agit d'un scalaire calculé sur le signal en volts, et non en pression, et donné en décibels, en prenant un référentiel de 1 Volt. Le niveau est obtenu en calculant, toutes les secondes, la moyenne quadratique du signal, et en moyennant sur la durée de la scène. Un filtrage de type A est opéré avant le calcul des moyennes quadratiques. Nous notons $L$, $L(E)$ et $L(T)$ les niveaux calculés en tenant compte respectivement de l'ensemble des samples, des samples d'événements et de textures.
% \end{itemize}

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of participants in experiment 1.b, we choose not to normalize the pleasantness scores.

\item \emph{semantic features}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for analysis. For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.

\item \emph{structural features}: features related to the \emph{sound level}. To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
\end{itemize}

@@
